---
title: "Group_1_Project"
author: "Matthew Runyon, Winston Wong, & Leo D'Agostino"
date: "10/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The data set that we have chosen to work with is the "Volcano Eruptions" data set, coming from Week 20 of the 2020 TidyTuesday project. This data is provided by the Smithsonian Institute. The data, which is split into five files (`volcano`, `eruptions`, `events`, `tree_rings`, and `sulfur`) and can be read in using the first half of the following code. Following that, we created the file paths to save various data sets, figures, and summary tables. Finally, we saved each of these files within the actual `Data_Sets` file on GitHub using the second half of code. As far as responsibilities for each of these files, Winston was responsible for data analysis on the `tree_rings` and `sulfur` data sets, Leo was responsible for investigating the `eruptions` data set, and Matthew was responsible for analyzing the `volcano` data set. Everyone provided assistance across data sets with code formatting to bolster the appearance after preliminary analyses had been completed independently. We chose not to delve into the `events` file because it was mainly date information without many other important variables to work with.

```{r}
volcano <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/volcano.csv')
eruptions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/eruptions.csv')
events <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/events.csv')
tree_rings <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/tree_rings.csv')
sulfur <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-05-12/sulfur.csv')

data_sets_path<-"C:/Users/mrunyon2/Documents/Classes/Fall 2020/CPSC 441/Group_1/Data_Sets"
figure_path<-"C:/Users/mrunyon2/Documents/Classes/Fall 2020/CPSC 441/Group_1/Figures"
summary_text_path<-"C:/Users/mrunyon2/Documents/Classes/Fall 2020/CPSC 441/Group_1/Summary_Text_files"

write.table(volcano, file.path(data_sets_path, "volcano.txt"))
write.table(eruptions, file.path(data_sets_path, "eruptions.txt"))
write.table(events, file.path(data_sets_path, "events.txt"))
write.table(tree_rings, file.path(data_sets_path, "tree_rings.txt"))
write.table(sulfur, file.path(data_sets_path, "sulfur.txt"))
```

__Loading Relevant Packages__

Here we load all of the relevant packages that will be used with our project. This ensures that all packages will be loaded at the start so that there is no redundancy within the code.
```{r}
library(tidyverse)
library(car)
library(ggpubr)
library(multcompView)
library(threejs)
library(qwraps2)
```

__Sulfur Concentration Dataset__

The sulfur concentration data set contains three variables. Two are sulfur content in ng/g collected from NEEM and WDC ice cores, which were collected from Greenland and Antarctica, respectively. These are treated as the response variables. The independent variable is the year, which is approximated as the time that the sulfur depositions took place.

___Summary Statistics and Means___

First, we are presenting summary statistics of this data. This is done by constructing a summary table based on the `qwraps2` package. Then, we save the means of the data set in a separate text file using the code below. Since there are only two numeric variables in this data set, this is not a large file on its own.
```{r}
sulfur_summary <-
  list("NEEM" =
       list("min_neem"       = ~ min(sulfur$neem, na.rm=T),
            "max_neem"       = ~ max(sulfur$neem, na.rm=T),
            "median_neem"    = ~ median(sulfur$neem, na.rm=T),
            "mean_neem"      = ~ mean(sulfur$neem, na.rm=T),
            "variance_neem"  = ~ var(sulfur$neem, na.rm=T),
            "stdev_neem"     = ~ sd(sulfur$neem, na.rm=T)),
       "WDC" =
       list("min_wdc"       = ~ min(sulfur$wdc, na.rm=T),
            "max_wdc"       = ~ max(sulfur$wdc, na.rm=T),
            "median_wdc"    = ~ median(sulfur$wdc, na.rm=T),
            "mean_wdc"      = ~ mean(sulfur$wdc, na.rm=T),
            "variance_wdc"  = ~ var(sulfur$wdc, na.rm=T),
            "stdev_wdc"     = ~ sd(sulfur$wdc, na.rm=T))
       )
sulfur_summary_table<-as.data.frame(summary_table(sulfur, summaries=sulfur_summary))
sulfur_summary_table

sulfur_means<-data.frame(mean(sulfur$neem, na.rm=T), mean(sulfur$wdc, na.rm=T))
names(sulfur_means)<-c("NEEM_mean", "WDC_mean")
row.names(sulfur_means)<-c("overall_mean_sulfur_content")
write.table(sulfur_means, file.path(summary_text_path, "sulfur_dataset_overall_means.txt"))
```

Plotted below is a simple line graph showing sulfur concentration in ice cores across years of deposition. We can see several major spikes in sulfur content, which are indicative of volcanic eruptions that have sent high-sulfur ash into the air. Notable peaks in the NEEM core data are around 525 AD, 540 AD, 575 AD, 625 AD, and 685 AD, while notable peaks in the WDC core data are around 550 AD, 575 AD, and 685 AD. For several of these major peaks, there appears to be overlap between the data sets. This is particularly interesting because these would signify major volcanic eruptions, as cores were pulled from ice samples near poles at opposite ends of the world.
```{r}
ggplot(sulfur, aes(x=year, y=neem))+
  geom_line()+
  ggtitle("Sulfur Concentration per Year from NEEM Ice Cores")+
  labs(x="Year (AD)", y="Sulfur Content (ng/g)")+
  theme(panel.grid=element_blank())+
  ggsave(file.path(figure_path, "neem_sulfur_across_years.pdf"))

ggplot(sulfur, aes(x=year, y=wdc))+
  geom_line()+
  ggtitle("Sulfur Concentration per Year from WDC Ice Cores")+
  labs(x="Year (AD)", y="Sulfur Content (ng/g)")+
  theme(panel.grid=element_blank())+
  ggsave(file.path(figure_path, "wdc_sulfur_across_years.pdf"))
```

___Linear Regression Analysis of Sulfur Content vs. Year and NEEM vs WDC Sulfur Content___

Here, we conduct linear regression to evaluate whether there is a significant linear relationship between year and sulfur content in either of the two locations.

First we will construct the model, then we will evaluate the assumptions to see if building that model is appropriate for this data set. Building a linear model allows us to establish a relationship between year and sulfur levels at NEEM and WDC locations in the form of sulfur levels as a function for year. Null and alternative hypotheses for each of the models are provided, followed by linear models themselves for sulfur levels at both locations are shown below. Slope and intercept values are provided upon calling the model.
```{r}
#Linear model for Sulfur levels at NEEM
#Ho: Slope = 0 (no linear relationship)
#HA: Slope not equal to 0 (linear relationship)
NEEM_linear_model <- lm(neem ~ year, data = sulfur)
NEEM_linear_model

#Linear model for Sulfur levels at WDC
#Ho: Slope = 0 (no linear relationship)
#HA: Slope not equal to 0 (linear relationship)
WDC_linear_model <- lm(wdc ~ year, data = sulfur) 
WDC_linear_model

#Linear model for NEEM vs. WDC
#Ho: Slope = 0 (no linear relationship)
#HA: Slope not equal to 0 (linear relationship)
NEEMvsWDC_linear_model <-lm(wdc~neem, data = sulfur)
NEEMvsWDC_linear_model
```

After forming these three linear models, we can to test for correlation between year vs. sulfur concentrations in both locations and correlation between NEEM and WDC sulfur content. We will  test for correlation using the Pearson's "r". Based off of these values, we can conclude that there is an extremely week linear correlation among the variables of interest.
```{r}
#Year vs NEEM
cor(sulfur$year, sulfur$neem, method = "pearson", use = "complete.obs")

#Year vs WDC
cor(sulfur$year, sulfur$wdc, method = "pearson", use = "complete.obs")

#NEEM vs WDC
cor(sulfur$neem, sulfur$wdc, method = "pearson", use = "complete.obs")
```

By running the `summary()` below function, we can determine the significance of our models. Our conclusions are noted as follows:

1) Year vs NEEM: p = 0.36. At alpha = 0.05, we fail to reject the null hypothesis and conclude that no significant linear relationship exists between year and NEEM sulfur content.

2) Year vs WDC: p = 0.0003. At alpha = 0.05, we reject the null hypothesis and conclude that there is a significant linear relationship between year and WDC sulfur content.

3) NEEM vs WDC: p = 3.2 x 10E-8. At alpha = 0.05, we reject the null hypothesis and conclude that there is a significant linear relationship between sulfur content at NEEM and WDC.
```{r}
#model summary for sulfur levels at NEEM
summary(NEEM_linear_model) 

#model summary for sulfur levels at WDC
summary(WDC_linear_model)  

#model summary for NEEM vs. WDC
summary(NEEMvsWDC_linear_model)
```

Now that we have constructed models and determined that a linear relationship was observed for two of the three models, we need to test the assumptions to determine whether linear regression was appropriate for this data. We can first create scatter plots to visualize a possible linear relationship between the independent and dependent variables. From the scatterplot it does not appear that there is a linear relationship between year and sulfur concentration at each location. Under normal circumstances, we would conclude that linear regression is inappropriate because there doesn't appear to be visual linear trend; nonetheless, for demonstration purposes, we are continuing with the process. We demonstrate the correlation coefficient and significance level on the charts using the `stat_cor()` function from the `ggpubr` package, as well as a confidence interval using the `geom_smooth()` function. We will continue to check the assumptions for demonstration purposes.
```{r}
ggplot(sulfur, aes(x=year, y=neem))+
  geom_point(alpha=1/5)+
  ggtitle("Year vs NEEM Sulfur Content (ng/g)")+
  labs(x="Year", y="NEEM Sulfur Content (ng/g)")+
  theme(panel.grid=element_blank())+
  stat_cor()+
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE)+
  ggsave(file.path(figure_path, "year_vs_neem_correlation.pdf"))

ggplot(sulfur, aes(x=year, y=wdc))+
  geom_point(alpha=1/5)+
  ggtitle("Year vs WDC Sulfur Content (ng/g)")+
  labs(x="Year", y="WDC Sulfur Content (ng/g)")+
  theme(panel.grid=element_blank())+
  stat_cor()+
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE)+
  ggsave(file.path(figure_path, "year_vs_wdc_correlation.pdf"))

ggplot(sulfur, aes(x=neem, y=wdc))+
  geom_point(alpha=1/5)+
  ggtitle("NEEM Sulfur Content (ng/g) vs WDC Sulfur Content (ng/g)")+
  labs(x="NEEM Sulfur Content (ng/g)", y="WDC Sulfur Content (ng/g)")+
  theme(panel.grid=element_blank())+
  stat_cor()+
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE)+
  ggsave(file.path(figure_path, "neem_vs_wdc_correlation.pdf"))

############################################

#Scatter plot for Year vs. NEEM
scatter_YearvNEEM <- ggplot(data = sulfur, aes(x = year, y= neem)) +
                          geom_point(color = "yellow") +
                          ylab("Sulfur detected in ng/g") +
                          ggtitle("Sulfur detected in ng/g from NEEM") 

#Scatter plot for Year vs. WDC
scatter_YearvWDC <- ggplot(data = sulfur, aes(x = year, y= wdc)) +
                          geom_point(color = "orange") +
                          ylab("Sulfur detected in ng/g") +
                          ggtitle("Sulfur detected in ng/g from WDC") 

#Scatter plot for NEEM vs WDC
scatter_NEEMvWDC <- ggplot(data = sulfur, aes(x = neem, y= wdc)) +
                          geom_point(color = "blue") +
                          ylab("WDC sulfur levels in ng/g") +
                          xlab("NEEM sulfur levels in ng/g")+
                          ggtitle("NEEM vs. WDC Sulfur Levels ") 
```

Next, we can create boxplots to check for outliers for the variables `neem` and `wdc` sulfur concentrations. From both boxplots, we conclude that there are many outliers, and that the data from both data sets is heavily right skewed and not normally distributed. If we look instead at the residuals from each of the models, we can see that the residuals are likewise heavily right skewed and not normally distributed. Again in this case, this would normally disqualify us from carrying out linear regression.
```{r}
sulfur_long<-pivot_longer(sulfur, cols=c("neem", "wdc"), names_to="core_loc")
sulfur_long$core_loc<-as.factor(sulfur_long$core_loc)
write.table(sulfur_long, file.path(summary_text_path, "sulfur_dataset_pivot_long.txt"))

#Boxplots for distribution of raw data
ggplot(sulfur_long, aes(x=year, y=value))+
  geom_boxplot(col="black", fill="yellow")+
  ggtitle("Boxplots of Sulfur Content in Ice Core Samples")+
  labs(y="Sulfur (ng/g)", x="")+
  theme(panel.grid=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())+
  facet_wrap(~core_loc)+
  ggsave(file.path(figure_path, "sulfur_distribution_boxplot.pdf"))

#Boxplots for distribution of residuals
neem_res<-as.data.frame(residuals(NEEM_linear_model))
wdc_res<-as.data.frame(residuals(WDC_linear_model))
neem_vs_wdc_res<-as.data.frame(residuals(NEEMvsWDC_linear_model))
sulfur_residuals<-cbind(neem_res, wdc_res, neem_vs_wdc_res)
sulfur_residuals_long<-pivot_longer(sulfur_residuals, cols=c("residuals(NEEM_linear_model)", "residuals(WDC_linear_model)", "residuals(NEEMvsWDC_linear_model)"), names_to="residual")

ggplot(sulfur_residuals_long, aes(y=value))+
  geom_boxplot(col="black", fill="orange")+
  ggtitle("Boxplots of Residuals for Sulfur Content Models")+
  labs(y="Residual Value", x="")+
  theme(panel.grid=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())+
  facet_wrap(~residual)+
  ggsave(file.path(figure_path, "sulfur_resiudals_distribution_boxplot.pdf"))

###################################
  
#Boxplot for sulfur levels from NEEM created using base R
sulfur$neem<-as.numeric(sulfur$neem)
boxplot(sulfur$neem, main = "Sulfur detected in ng/g from NEEM",
        xlab = "sulfur (ng/g)", horizontal = TRUE)

#Boxplot for sulfur levels from NEEM created using ggplot2
ggplot(sulfur, aes(y=neem))+
  geom_boxplot()+
  ggtitle("Boxplot of NEEM Sulfur Concentrations")+
  labs(y="NEEM Sulfur Concentration (ng/g)")+
  theme(panel.grid=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())

#Boxplot for sulfur levels from WDC created using base R
WDCsulfur <- as.numeric(sulfur$wdc)
boxplot(sulfur$wdc, main = "Sulfur detected in ng/g from WDC",
        xlab = "sulfur (ng/g)", horizontal = TRUE)

#Boxplot for sulfur levels from WDC created using ggplot2
ggplot(sulfur, aes(y=wdc))+
  geom_boxplot()+
  ggtitle("Boxplot of WDC Sulfur Concentrations")+
  labs(y="WDC Sulfur Concentration (ng/g)")+
  theme(panel.grid=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())
```

Overall, even though we demonstrated that the Year vs WDC and NEEM vs WDC models were statistically significant, we have shown through testing the assumptions that linear regression is not an appropriate approach to take with this data set due to extreme non-normality and lack of a suggestive visible linear trend. Therefore, if we were to look at these variables in a more detailed approach, an approach other than linear regression should be taken to test for relationship between these variables.

__Tree Ring Dataset__

Like the sulfur dataset, the tree ring data set contains three variables. Two response variables are the European temperature index and tree ring Z scores. The European temperature index is a relative temperature index in degrees Celsius. The more ash that is spewed into the air from volcanic eruptions, the more sunlight should be reflected in the atmosphere, and the cooler the average temperature for a given year should be. Likewise, temperature and seasonal extremes should cause more variability in vegetative growth such as tree rings, and this variability should be reflected by larger Z scores. The independent variable is the year from which these measurements were taken, dating back 2,000 years.

___Summary Statistics and Means___

First, we present summary statistics of this data. As done with the sulfur data set, this is done by constructing a summary table based on the `qwraps2` package. Then, we save the means of the data set in a separate text file using the code below. Again, there are only two numeric values in this data file.
```{r}
tree_ring_summary <-
  list("n_tree" =
         list("min_n_tree"       = ~ min(tree_rings$n_tree, na.rm=T),
              "max_n_tree"       = ~ max(tree_rings$n_tree, na.rm=T),
              "median_n_tree"    = ~ median(tree_rings$n_tree, na.rm=T),
              "mean_n_tree"      = ~ mean(tree_rings$n_tree, na.rm=T),
              "variance_n_tree"  = ~ var(tree_rings$n_tree, na.rm=T),
              "stdev_n_tree"     = ~ sd(tree_rings$n_tree, na.rm=T)),
       "europe_temp_index" =
         list("min_europe_temp_index"       = ~ min(tree_rings$europe_temp_index, na.rm=T),
              "max_europe_temp_index"       = ~ max(tree_rings$europe_temp_index, na.rm=T),
              "median_europe_temp_index"    = ~ median(tree_rings$europe_temp_index, na.rm=T),
              "mean_europe_temp_index"      = ~ mean(tree_rings$europe_temp_index, na.rm=T),
              "variance_europe_temp_index"  = ~ var(tree_rings$europe_temp_index, na.rm=T),
              "stdev_europe_temp_index"     = ~ sd(tree_rings$europe_temp_index, na.rm=T))
  )
tree_ring_summary_table<-as.data.frame(summary_table(tree_rings, summaries=tree_ring_summary))
tree_ring_summary_table

tree_ring_means<-data.frame(mean(tree_rings$n_tree, na.rm=T), mean(tree_rings$europe_temp_index, na.rm=T))
names(tree_ring_means)<-c("n_tree_mean", "europe_temp_index_mean")
row.names(tree_ring_means)<-c("overall_means")
write.table(tree_ring_means, file.path(summary_text_path, "tree_ring_dataset_overall_means.txt"))
```
Plotted below are two line graphs showing the measurements of tree ring Z-scores and the European temperature index over time. As shown in these two graphs, the data is very noisy, with large swings from year to year. It is much more noisy than the ice core data. This likely is due to the imprecision of these measurements; since these date back before records were being taken and the data isn't preserved in time like it is for the sulfur in the ice cores, there is a lot more room for error from these estimations.
```{r}
ggplot(tree_rings, aes(x=year, y=n_tree))+
  geom_line()+
  ggtitle("Tree Ring Z-Scores per Year")+
  labs(x="Year (AD)", y="Z-Score")+
  theme(panel.grid=element_blank())+
  ggsave(file.path(figure_path, "tree_ring_scores_across_years.pdf"))

ggplot(tree_rings, aes(x=year, y=europe_temp_index))+
  geom_line()+
  ggtitle("European Temperature Index per Year")+
  labs(x="Year (AD)", y="European Temperature Index")+
  theme(panel.grid=element_blank())+
  ggsave(file.path(figure_path, "temp_index_across_years.pdf"))

```

___Linear Regression Analysis of Sulfur Content vs. Year and NEEM vs WDC Sulfur Content___

Here, we conduct linear regression to evaluate whether there is a significant linear relationship between year and tree ring Z-score, year and temperature index, and tree ring Z-score and temperature index.

First we will construct the representative models, then we will evaluate the assumptions to see if building that model is appropriate for this data set. Null and alternative hypotheses for each of the models are provided, followed by linear models themselves are shown below. Slope and intercept values are provided upon calling the model.

```{r}
#Linear model for year vs "n_tree"
#Ho: Slope = 0 (no linear relationship)
#HA: Slope not equal to 0 (linear relationship)
nTree_linear_model <- lm(n_tree ~ year, data = tree_rings)
nTree_linear_model

#Linear model for year vs. "europe_temp_index"
#Ho: Slope = 0 (no linear relationship)
#HA: Slope not equal to 0 (linear relationship)
ETI_linear_model <- lm(europe_temp_index ~ year, data = tree_rings) 
ETI_linear_model

#Linear model for "n_tree" vs. "europe_temp_index
#Ho: Slope = 0 (no linear relationship)
#HA: Slope not equal to 0 (linear relationship)
nTreevsETI_linear_model <-lm(europe_temp_index~n_tree, data = tree_rings)
nTreevsETI_linear_model
```

After forming these three linear models, we can to test for correlation between year vs. sulfur concentrations in both locations and correlation between NEEM and WDC sulfur content. We will  test for correlation using the Pearson's "r". Based off of these values, we can conclude that there is a weak negative relationship between year and tree ring Z-score and between year and European temperature index, and a strong positive linear correlation between tree ring Z-score and European temperature index.
```{r}
#Year vs n_tree
cor(tree_rings$year, tree_rings$n_tree, method = "pearson", use = "complete.obs")

#Year vs europe_temp_index
cor(tree_rings$year, tree_rings$europe_temp_index, method = "pearson", use = "complete.obs")

#n_tree vs europe_temp_index
cor(tree_rings$n_tree, tree_rings$europe_temp_index, method = "pearson", use = "complete.obs")
```

By running the `summary()` below function, we can determine the significance of our models. Our conclusions are noted as follows:

1) Year vs n_tree: p = 2.6E-11. At alpha = 0.05, we reject the null hypothesis and conclude that there is a significant linear relationship between year and tree ring Z-score.

2) Year vs europe_temp_index: p = 2.2E-16. At alpha = 0.05, we reject the null hypothesis and conclude that there is a significant linear relationship between year and European temperature index.

3) n_tree vs europe_temp_index: p = 2.2E-16. At alpha = 0.05, we reject the null hypothesis and conclude that there is a significant linear relationship between tree ring Z-score and European temperature index.
```{r}
#model summary for "n_tree" and year
summary(nTree_linear_model) 

#model summary for "europe_temp_index" and year
summary(ETI_linear_model)  

#model summary for "n_tree" vs. "europe_temp_index"
summary(nTreevsETI_linear_model)
```

Now that we have constructed models and determined that a linear relationship was observed for two of the three models, we need to test the assumptions to determine whether linear regression was appropriate for this data. We can first create scatter plots to visualize a possible linear relationship between the independent and dependent variables. From the scatterplot it is possible that there is a linear relationship between year and tree ring Z-score or year and European temperature index. However, it appears that there is a strong, positive linear relationship between tree ring Z-score and European temperature index, as was suggested by the linear model. Because a possible linear trend exists for all three sets of data, we will continue checking the model assumptions.

```{r}
ggplot(tree_rings, aes(x=year, y=n_tree))+
  geom_point(alpha=1/5)+
  ggtitle("Year vs Tree Ring Z-Score")+
  labs(x="Year", y="Tree Ring Z-Score")+
  theme(panel.grid=element_blank())+
  stat_cor()+
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE)+
  ggsave(file.path(figure_path, "year_vs_tree_ring_correlation.pdf"))

ggplot(tree_rings, aes(x=year, y=europe_temp_index))+
  geom_point(alpha=1/5)+
  ggtitle("Year vs European Temperature Index")+
  labs(x="Year", y="European Temperature Index")+
  theme(panel.grid=element_blank())+
  stat_cor()+
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE)+
  ggsave(file.path(figure_path, "year_vs_temp_index_correlation.pdf"))

ggplot(tree_rings, aes(x=n_tree, y=europe_temp_index))+
  geom_point(alpha=1/5)+
  ggtitle("Tree Ring Z-Score vs European Temperature Index")+
  labs(x="Tree Ring Z-Score", y="European Temperature Index")+
  theme(panel.grid=element_blank())+
  stat_cor()+
  geom_smooth(method=lm , color="red", fill="#69b3a2", se=TRUE)+
  ggsave(file.path(figure_path, "tree_ring_vs_temp_index_correlation.pdf"))

###########################################

#Scatter plot for Year vs. "n_Tree"
scatter_YearvnTree <- ggplot(data = tree_rings, aes(x = year, y= n_tree)) +
                          geom_point(color = "green") +
                          ylab("Tree ring z-scores relative to year = 1000-1099") +
                          ggtitle("Tree rings z-scores vs. Year") 

#Scatter plot for Year vs. "europe_temp_index"
scatter_YearvETI <- ggplot(data = tree_rings, aes(x = year, y= europe_temp_index)) +
                          geom_point(color = "red") +
                          ylab("Pages 2K Temperature for Europe in Celsius") +
                          ggtitle("Pages 2K Temperature for Europe in Celsius vs. Year") 

#Scatter plot for "n_tree" vs "europe_temp_index"
scatter_nTreevETI <- ggplot(data = tree_rings, aes(x = n_tree, y= europe_temp_index)) +
                          geom_point(color = "blue") +
                          ylab("Pages 2K Temperature for Europe in Celsius") +
                          xlab("Tree ring z-scores relative to year = 1000-1099")+
                          ggtitle("Tree ring z-scores vs. ETI in Celsius ") 
```

Next, we can create boxplots to check for outliers for the variables `n_tree` and `europe_temp_index` values. From both boxplots, we conclude that there are several outliers, but the data from both data appears normally distributed without skew. If we look instead at the residuals from each of the models, they also appear to follow a normal distribution for all three of the models. From this data, we can conclude that this assumption has been met.
```{r}
tree_rings_long<-pivot_longer(tree_rings, cols=c("n_tree", "europe_temp_index"), names_to="measurement_type")
tree_rings_long$measurement_type<-as.factor(tree_rings_long$measurement_type)
write.table(tree_rings_long, file.path(summary_text_path, "tree_rings_dataset_pivot_long.txt"))

#Boxplots for distribution of raw data
ggplot(tree_rings_long, aes(x=year, y=value))+
  geom_boxplot(col="black", fill="blue")+
  ggtitle("Boxplots of Tree Ring Z-Scores and European Temperature Indices")+
  labs(y="Value", x="")+
  theme(panel.grid=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())+
  facet_wrap(~measurement_type)+
  ggsave(file.path(figure_path, "tree_ring_temp_index_distribution_boxplot.pdf"))

#Boxplots for distribution of residuals
n_tree_res<-as.data.frame(residuals(nTree_linear_model))
ETI_res<-as.data.frame(residuals(ETI_linear_model))
n_tree_ETI_res<-as.data.frame(residuals(nTreevsETI_linear_model))
tree_ring_residuals<-cbind(n_tree_res, ETI_res, n_tree_ETI_res)
tree_ring_residuals_long<-pivot_longer(tree_ring_residuals, cols=c("residuals(nTree_linear_model)", "residuals(ETI_linear_model)", "residuals(nTreevsETI_linear_model)"), names_to="residual")

ggplot(tree_ring_residuals_long, aes(y=value))+
  geom_boxplot(col="black", fill="green")+
  ggtitle("Boxplots of Residuals for Tree Ring Dataset Models")+
  labs(y="Residual Value", x="")+
  theme(panel.grid=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())+
  facet_wrap(~residual)+
  ggsave(file.path(figure_path, "tree_ring_temp_index_residuals_distribution_boxplot.pdf"))

#############################

#Boxplot for "n_tree"
bp_nTree<- ggplot(tree_rings, aes(y=n_tree))+
  geom_boxplot(fill = "green")+
  ggtitle("Boxplot of Tree ring z-scores relative to year = 1000-1099") +
  labs(y="Tree ring z-scores ")+
  theme(panel.grid=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())

#Boxplot for "europe_temp_index"
bp_ETI <- ggplot(tree_rings, aes(y=europe_temp_index))+
  geom_boxplot(fill = "red")+
  ggtitle("Pages 2K Temperature for Europe in Celsius") +
  labs(y="Celsius ")+
  theme(panel.grid=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())
```

Overall, we have demonstrated that the assumptions for conducting linear regression have been met, so this form of analysis is appropriate. Based off of these models, we can conclude that there is a weak negative linear relationship between year and tree ring Z-score and between year and European temperature index, and a strong positive linear relationship between tree ring Z-score and European temperature index. As noisy as the data was from the initial line plot, these conclusions were rather interesting, as we wouldn't have thought that they would have lined up so well. However, it does make sense that vegetative growth of trees would be influenced by temperature, so this correlation is reasonable.

__Volcano Dataset__

The volcano data set contains multiple variables related to characteristics of known volcanos around the world. Some variables specify volcano location, such as `country`, `region`,`subregion`, `latitude`, and `longitude`. Other variables specify features of the volcano, such as `primary_volcano_type`, `tectonic_settings`, `elevation`, `last_eruption_year`, and various kinds of rock types that are characteristic of that volcano. Finally, human populations within 5, 10, 30, and 100 km of each volcano are provided.

___Data Cleanup___

First I cleaned up the data in the `primary_volcano_type` column by removing all duplicate values of plural classifiers (removed all (es) and (s)). Then I coverted this column, as well as `country`, `region`, `subregion`, `tectonic_settings`, `evidence_category`, `major_rock_.`, and `minor_rock_.` to factors. I also changed all the unknown `last_eruption_year` data with `NA` and converted those to numeric values. Finally, I arrange the data, create a new columns for years since the last eruption date using the `last_eruption_year` variable, and order the columns how I want them.


```{r}
volcano$primary_volcano_type<-gsub(".(es).", "", volcano$primary_volcano_type)
volcano$primary_volcano_type<-gsub(".(s).", "", volcano$primary_volcano_type)
volcano$primary_volcano_type<-sub("Stratovolcano.", "Stratovolcano", volcano$primary_volcano_type)
volcano$primary_volcano_type<-as.factor(volcano$primary_volcano_type)
volcano$country<-as.factor(volcano$country)
volcano$region<-as.factor(volcano$region)
volcano$subregion<-as.factor(volcano$subregion)
volcano$tectonic_settings<-as.factor(volcano$tectonic_settings)
volcano$evidence_category<-as.factor(volcano$evidence_category)
volcano$major_rock_1<-as.factor(volcano$major_rock_1)
volcano$major_rock_2<-as.factor(volcano$major_rock_2)
volcano$major_rock_3<-as.factor(volcano$major_rock_3)
volcano$major_rock_4<-as.factor(volcano$major_rock_4)
volcano$major_rock_5<-as.factor(volcano$major_rock_5)
volcano$minor_rock_1<-as.factor(volcano$minor_rock_1)
volcano$minor_rock_2<-as.factor(volcano$minor_rock_2)
volcano$minor_rock_3<-as.factor(volcano$minor_rock_3)
volcano$minor_rock_4<-as.factor(volcano$minor_rock_4)
volcano$minor_rock_5<-as.factor(volcano$minor_rock_5)
volcano$last_eruption_year<-as.numeric(gsub("Unknown", NA, volcano$last_eruption_year))
volcano<-volcano%>%
  arrange(primary_volcano_type, last_eruption_year, elevation)%>%
  mutate(years_ago=2020-last_eruption_year)
ordered_columns<-c("volcano_number", "volcano_name", "region", "subregion", "country", "latitude", "longitude", "tectonic_settings", "primary_volcano_type", "elevation", "last_eruption_year", "years_ago", "evidence_category", "population_within_5_km","population_within_10_km","population_within_30_km","population_within_100_km", "major_rock_1","major_rock_2","major_rock_3","major_rock_4","major_rock_5","minor_rock_1","minor_rock_2","minor_rock_3","minor_rock_4","minor_rock_5")
volcano<-volcano[ordered_columns]
volcano
write.table(volcano, file.path(data_sets_path,"updated_volcano_data_set.txt"))
```

___Summary Statistics___

Here I compile a table of all the means of relevant numeric data from the `volcano` data set. Out of the 27 variables in this data file, I selected `elevation`, `last_eruption_year`, `years_ago`, `population_within_5_km`, `population_within_10_km`,`population_within_30_km`, and `population_within_100_km`. I first looked at overall column means for these variables and saved a file of overall means titled `"volcano_dataset_overall_means.txt"`, then I looked at means based on the `primary_volcano_type` variable. I rounded each of these to whole units. Finally, I saved a file with the means for each variable by primary volcano type, titled `"volcano_dataset_means_by_primary_volcano_type.txt"`.
```{r}
overall_means<-data.frame(mean(volcano$elevation),
mean(volcano$last_eruption_year, na.rm=T),
mean(volcano$years_ago, na.rm=T),
mean(volcano$population_within_5_km, na.rm=T),
mean(volcano$population_within_10_km, na.rm=T),
mean(volcano$population_within_30_km, na.rm=T),
mean(volcano$population_within_100_km, na.rm=T))
names(overall_means)<-c("elevation", "last_eruption_year", "years_ago", "population_within_5_km", "population_within_10_km","population_within_30_km", "population_within_100_km")
row.names(overall_means)<-c("overall_means")
write.table(overall_means, file.path(summary_text_path, "volcano_dataset_overall_means.txt"))

elevation_means<-volcano%>%
  group_by(primary_volcano_type)%>%
  summarise(mean_height = mean(elevation))
elevation_means$mean_height<-round(elevation_means$mean_height,0)

last_eruption_year_means<-volcano%>%
  group_by(primary_volcano_type)%>%
  summarise(mean_year = mean(last_eruption_year, na.rm=T))
last_eruption_year_means$mean_year<-round(last_eruption_year_means$mean_year,0)

years_ago<-volcano%>%
  group_by(primary_volcano_type)%>%
  summarise(mean_years_ago=mean(years_ago, na.rm=T))
years_ago$mean_years_ago<-round(years_ago$mean_years_ago, 0)

five_km_pop_means<-volcano%>%
  group_by(primary_volcano_type)%>%
  summarise(mean_5km_pop=mean(population_within_5_km, na.rm=T))
five_km_pop_means$mean_5km_pop<-round(five_km_pop_means$mean_5km_pop, 0)

ten_km_pop_means<-volcano%>%
  group_by(primary_volcano_type)%>%
  summarise(mean_10km_pop=mean(population_within_10_km, na.rm=T))
ten_km_pop_means$mean_10km_pop<-round(ten_km_pop_means$mean_10km_pop, 0)

thirty_km_pop_means<-volcano%>%
  group_by(primary_volcano_type)%>%
  summarise(mean_30km_pop=mean(population_within_30_km, na.rm=T))
thirty_km_pop_means$mean_30km_pop<-round(thirty_km_pop_means$mean_30km_pop, 0)

hundred_km_pop_means<-volcano%>%
  group_by(primary_volcano_type)%>%
  summarise(mean_100km_pop=mean(population_within_100_km, na.rm=T))
hundred_km_pop_means$mean_100km_pop<-round(hundred_km_pop_means$mean_100km_pop, 0)

volcano_means_table <- merge(merge(merge(merge(merge(merge(
  elevation_means,
  last_eruption_year_means, all=T),
  years_ago, all=T),
  five_km_pop_means, all=T),
  ten_km_pop_means, all=T),
  thirty_km_pop_means, all=T),
  hundred_km_pop_means, all=T)
write.table(volcano_means_table, file.path(summary_text_path, "volcano_dataset_means_by_primary_volcano_type.txt"))
```

___Types of Volcanoes___

Here I create a bar chart to show the types of volcanos present among the provided data. Based on this chart, we can see that, by a large margin, stratovolcanos are the most common type of volcano.

```{r}
type_count<-volcano %>% 
  group_by(primary_volcano_type) %>%
  summarise(no_rows1 = length(primary_volcano_type))

ggplot(volcano, aes(x=primary_volcano_type))+
  geom_bar()+
  ggtitle("Number of Volcanos by Type")+
  labs(x="Type of Volcano", y="Number of Volcanos")+
  theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.95))+
  geom_text(data=type_count, aes(x=primary_volcano_type, y=no_rows1, label=no_rows1), nudge_y=30)+
  ggsave(file.path(figure_path,"volcano_count_by_type.pdf"))
```

Here I create a bar chart showing the number of volcanos in each tectonic subcategory. Here we can see that the most common tectonic setting is for subduction zone / continental crust.

```{r}
tectonic_count<-volcano %>% 
  group_by(tectonic_settings) %>%
  summarise(no_rows3 = length(tectonic_settings))

ggplot(volcano, aes(x=tectonic_settings))+
  geom_bar()+
  ggtitle("Number of Volcanos by Tectonic Setting")+
  labs(x="Tectonic Setting", y="Number of Volcanos")+
  theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.95))+
  geom_text(data=tectonic_count ,aes(x=tectonic_settings, y=no_rows3, label=no_rows3), nudge_y=50)+
  scale_x_discrete(labels = function(tectonic_settings) str_wrap(tectonic_settings, width = 20))+
   ggsave(file.path(figure_path, "volcano_count_by_tectonic_setting.pdf"))
```

___Elevation___

Here is a histogram showing the frequency of different volcano heights. The majority of volcanos are under 2,500 feet tall, with most being 1,000 to 1,500 feet tall. The data do appear to be moderately skewed to the right.
```{r}
ggplot(volcano, aes(x=elevation))+
  geom_histogram(binwidth=500, color="black")+
  ggtitle("Histogram of Volcano Elevations")+
  labs(x="Elevation", y="Observations")+
  theme(panel.grid=element_blank())+
  ggsave(file.path(figure_path, "volcano_elevation_histogram.pdf"))
```

___Means and Multiple Comparisons of Elevation by Volcano Type___

Here I am using the `car` package for the Levene's Test of Homogeneity of Variance. Normal distribution of residuals is evaluated using the `shapiro.test()` function. Under normal circumstances, homogeneity of variance and normality of residuals would be considered prerequisite assumptions for running an ANOVA; however, for all data, we treat the data as if the assumptions have been met, even if they are demonstrably not.
```{r}
mean(volcano$elevation)

ggplot(volcano, aes(x=elevation, y=primary_volcano_type))+
  geom_boxplot()+
  ggtitle("Boxplot of Volcano Elevation by Volcano Type")+
  labs(x="Elevation (Meters above Sea Level)", y="Volcano Type")+
  coord_flip()+
  theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.95))+
  stat_summary(fun=mean, geom="point", pch="+", size=2.5, color="red", fill="red")+
  ggsave(file.path(figure_path, "elevation_by_type_boxplot.pdf"))


#Ho: All elevations are equal among volcano types.
#HA: At least one mean elevation is different.
model1=lm(elevation~primary_volcano_type, data=volcano)
res_values1<-residuals(model1)
shapiro.test(res_values1)
leveneTest(model1)
#Here I am doing an ANOVA even though normal distribution of residuals and homogeneity of variance are not met.
elevation_anova<-aov(model1)
summary(elevation_anova)
```

___Multiple Comparisons (Volcano Type)  of Population - 5km___
```{r}
ggplot(volcano, aes(x=population_within_5_km, y=primary_volcano_type))+
  geom_boxplot()+
  ggtitle("Boxplot of Population within 5 km by Volcano Type")+
  labs(x="Population", y="Volcano Type")+
  coord_flip()+
  theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.95))+
  stat_summary(fun=mean, geom="point", pch="+", size=2.5, color="red", fill="red")+
  ggsave(file.path(figure_path, "5km_pop_by_type_boxplot.pdf"))

#Ho: All populations are equal among volcano types
#HA: At least one mean population is different
model2=lm(population_within_5_km~primary_volcano_type, data=volcano)
res_values2<-residuals(model2)
shapiro.test(res_values2)
leveneTest(model2)
five_km_anova<-aov(model2)
summary(five_km_anova)
```

___Multiple Comparisons of (Volcano Type) Population - 10 km___
```{r}
ggplot(volcano, aes(x=population_within_10_km, y=primary_volcano_type))+
  geom_boxplot()+
  ggtitle("Boxplot of Population within 10 km by Volcano Type")+
  labs(x="Population", y="Volcano Type")+
  coord_flip()+
  theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.95))+
  stat_summary(fun=mean, geom="point", pch="+", size=2.5, color="red", fill="red")+
  ggsave(file.path(figure_path, "10km_pop_by_type_boxplot.pdf"))

#Ho: All populations are equal among volcano types
#HA: At least one mean population is different
model3=lm(population_within_10_km~primary_volcano_type, data=volcano)
res_values3<-residuals(model3)
shapiro.test(res_values3)
leveneTest(model3)
ten_km_anova<-aov(model3)
summary(ten_km_anova)
```

___Multiple Comparisons (Volcano Type) of Population - 30 km___
```{r}
ggplot(volcano, aes(x=population_within_30_km, y=primary_volcano_type))+
  geom_boxplot()+
  ggtitle("Boxplot of Population within 30 km by Volcano Type")+
  labs(x="Population", y="Volcano Type")+
  coord_flip()+
  theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.95))+
  stat_summary(fun=mean, geom="point", pch="+", size=2.5, color="red", fill="red")+
  ggsave(file.path(figure_path, "30km_pop_by_type_boxplot.pdf"))

#Ho: All populations are equal among volcano types
#HA: At least one mean population is different
model4=lm(population_within_30_km~primary_volcano_type, data=volcano)
res_values4<-residuals(model4)
shapiro.test(res_values4)
leveneTest(model4)
thirty_km_anova<-aov(model4)
summary(thirty_km_anova)
```

___Multiple Comparisons (Volcano Type) of Population - 100 km___
```{r}
ggplot(volcano, aes(x=population_within_100_km, y=primary_volcano_type))+
  geom_boxplot()+
  ggtitle("Boxplot of Population within 100 km by Volcano Type")+
  labs(x="Population", y="Volcano Type")+
  coord_flip()+
  theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.95))+
  stat_summary(fun=mean, geom="point", pch="+", size=2.5, color="red", fill="red")+
  ggsave(file.path(figure_path, "100km_pop_by_type_boxplot.pdf"))

#Ho: All populations are equal among volcano types
#HA: At least one mean population is different
model5=lm(population_within_100_km~primary_volcano_type, data=volcano)
res_values5<-residuals(model5)
shapiro.test(res_values5)
leveneTest(model5)
hundred_km_anova<-aov(model5)
summary(hundred_km_anova)
```

Next, I am showing essentially the same thing as above. I calculated the mean population per volcano type and plotted them in bar graphs. This is a better representation than the boxplots above because the outliers skewed the range and caused the boxes of the boxplot to be so compressed that no real differences could be seen. Plot area could have been adjusted in `ggplot()` to cut off these outliers, but I didn't feel as if that would be truly representative of the data. A word of caution for the bar plots that follow, because the outliers were so severe, they do skew the mean values reported.
```{r, figures-side, fig.show="hold"}
population_means_by_volcano_type<-group_by(volcano, primary_volcano_type) %>% 
  summarize(mean_population_within_5_km = mean(population_within_5_km), mean_population_within_10_km = mean(population_within_10_km), mean_population_within_30_km = mean(population_within_30_km), mean_population_within_100_km = mean(population_within_100_km))
population_means_by_volcano_type[,2:5]<-round(population_means_by_volcano_type[,2:5],digits=0)

ggplot(population_means_by_volcano_type, aes(x=primary_volcano_type, y=mean_population_within_5_km))+
  geom_col()+
  ggtitle("Mean Population Living within 5km of Volcanos by Volcano Type")+
  labs(x="Volcano Type", y="Mean Population within 5km")+
  theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.95))+
  geom_text(data=population_means_by_volcano_type, aes(x=primary_volcano_type, y=mean_population_within_5_km, label=mean_population_within_5_km), nudge_y=10000, size=3)+
  ggsave(file.path(figure_path, "mean_5km_pop_by_type_barplot.pdf"))

ggplot(population_means_by_volcano_type, aes(x=primary_volcano_type, y=mean_population_within_10_km))+
  geom_col()+
  ggtitle("Mean Population Living within 10km of Volcanos by Volcano Type")+
  labs(x="Volcano Type", y="Mean Population within 10km")+
  theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.95))+
  geom_text(data=population_means_by_volcano_type, aes(x=primary_volcano_type, y=mean_population_within_10_km, label=mean_population_within_10_km), nudge_y=10000, size=3)+
  ggsave(file.path(figure_path, "mean_10km_pop_by_type_barplot.pdf"))

ggplot(population_means_by_volcano_type, aes(x=primary_volcano_type, y=mean_population_within_30_km))+
  geom_col()+
  ggtitle("Mean Population Living within 30km of Volcanos by Volcano Type")+
  labs(x="Volcano Type", y="Mean Population within 30km")+
  theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.95))+
  geom_text(data=population_means_by_volcano_type, aes(x=primary_volcano_type, y=mean_population_within_30_km, label=mean_population_within_30_km), nudge_y=25000, size=3)+
  ggsave(file.path(figure_path, "mean_30km_pop_by_type_barplot.pdf"))

ggplot(population_means_by_volcano_type, aes(x=primary_volcano_type, y=mean_population_within_100_km))+
  geom_col()+
  ggtitle("Mean Population Living within 100km of Volcanos by Volcano Type")+
  labs(x="Volcano Type", y="Mean Population within 100km")+
  theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.95))+
  geom_text(data=population_means_by_volcano_type, aes(x=primary_volcano_type, y=mean_population_within_100_km, label=mean_population_within_100_km), nudge_y=175000, size=3)+
ggsave(file.path(figure_path, "mean_100km_pop_by_type_barplot.pdf"))
```

___Eruption Dates___

This is a histogram showing the number of volcanos that have a known eruption within a certian number of years in the past. Currently, the bin width is set at 1,000 years. However, this doesn't discriminate well for the most recent years. Something needs to be done to split up the more recent and the older data.
```{r}
ggplot(volcano, aes(x=last_eruption_year))+
  geom_histogram(binwidth=1000, color="black")+
  ggtitle("Histogram of Last Known Eruptions")+
  labs(x="Date", y="Observations")+
  theme(panel.grid=element_blank())+
  ggsave(file.path(figure_path, "last_eruption_year_histogram.pdf"))
```

___Multiple Comparisons - Last Known Eruption___

This analysis seeks to determine whether or not certain types of volcanoes are more likely to erupt than others based on their last known eruption date. In this boxplot, the further down the y-axis, the longer it has been since an eruption.
```{r}
ggplot(volcano, aes(x=last_eruption_year, y=primary_volcano_type))+
  geom_boxplot()+
  ggtitle("Boxplot of Last Eruption Year by Volcano Type")+
  labs(x="Year", y="Volcano Type")+
  coord_flip()+
  theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.95))+
  stat_summary(fun=mean, geom="point", pch="+", size=2.5, color="red", fill="red")+
   ggsave(file.path(figure_path, "eruption_year_by_type_boxplot.pdf"))

#Ho: Mean last eruption year is equal among all volcano types
#HA: At least one mean eruption year is not equal
model6=lm(last_eruption_year~primary_volcano_type, data=volcano)
res_values6<-residuals(model6)
shapiro.test(res_values6)
leveneTest(model6)
eruption_year_anova<-aov(model6)
summary(eruption_year_anova)
```

Much like above, this analysis seeks to determine if volcanos in different tectonic settings, rather than just different volcano types, are more likely than others to erupt more frequently. Here again, the rationale is that the further down the y-axis, the longer it has been since an eruption.
```{r}
ggplot(volcano, aes(x=last_eruption_year, y=tectonic_settings))+
  geom_boxplot()+
  ggtitle("Boxplot of Last Eruption Year by Tectonic Setting")+
  labs(x="Year", y="Tectonic Setting")+
  coord_flip()+
  theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.95))+
  stat_summary(fun=mean, geom="point", pch="+", size=2.5, color="red", fill="red")+
   ggsave(file.path(figure_path, "last_eruption_year_by_tectonic_setting.pdf"))

#Ho: Mean last eruption year is equal among all tectonic settings.
#HA: At least one mean eruption year is not equal.
model7=lm(last_eruption_year~tectonic_settings, data=volcano)
res_values7<-residuals(model7)
shapiro.test(res_values7)
leveneTest(model7)
eruption_tectonic_anova<-aov(model7)
summary(eruption_tectonic_anova)
```

___Linear Regression and Correlation of Volcano Elevation with Last Known Eruption Year___

Here, a reasonable assumption that could be made about volcanos is that ones that erupt more frequently are taller because they are adding more layers of molten rock to their outer surface. To test this assumption, I performed linear regression analysis between last known eruption year and elevation. Based on the results, we can conclude that there is no significant linear relationship between elevation and last known eruption date.
```{r}
#Ho: Slope = 0 (no linear relationship)
#HA: Slope not equal to 0 (linear relationship)

ggplot(volcano, aes(x=last_eruption_year, y=elevation))+
  geom_point(alpha=1/5)+
  ggtitle("Last Eruption Year vs Elevation")+
  labs(x="Year", y="Elevation")+
  theme(panel.grid=element_blank())+
  stat_cor()+
  ggsave(file.path(figure_path, "elevation_vs_eruption_year_correlation.pdf"))

```


___Global Volcano Locations___

Here I create a bar chart showing the number of volcanos present in various regions of the world. Since actual latitude and longitude coordinates were provided in this data file, following this regional grouping bar chart, I use the `threejs` package to plot each of the coordinates on a global map. Height of individual bars on the map represent volcano elevation, while heatmap colors indicate the number of years since the last known eruption (red = most recent, white = most distant past).
```{r}
region_count<-volcano %>% 
  group_by(region) %>%
  summarise(no_rows2 = length(region))

ggplot(volcano, aes(x=region))+
  geom_bar()+
  ggtitle("Number of Volcanos by Region")+
  labs(x="Region", y="Number of Volcanos")+
  theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.95))+
  geom_text(data=region_count ,aes(x=region, y=no_rows2, label=no_rows2), nudge_y=10)+
  ggsave(file.path(figure_path, "volcano_count_by_region.pdf"))

library(threejs)

for_globe<-volcano[,c("latitude", "longitude", "years_ago", "elevation")]
colors=as.numeric(cut(for_globe$years_ago,
                      breaks=1000, include.lowest=T, na.rm=T))
colors=heat.colors(colors)
globejs(lat=for_globe$latitude, long=for_globe$longitude, val=(0.025*for_globe$elevation), color=colors, atmosphere=T, pointsize=0.75, emissive=T)

#Might do something with the latitude groupings (Antarctic Circle, Tropic of Capricorn, Equator, Tropic of Cancer, Arctic Circle)
s<-volcano%>%filter(latitude >=-90 & latitude <=-66.5)
s2<-volcano%>%filter(latitude >=-66.5 & latitude <=-23.5)
s3<-volcano%>%filter(latitude >=-23.5 & latitude <=0)
s4<-volcano%>%filter(latitude >=0 & latitude <=23.5)
s5<-volcano%>%filter(latitude >=23.5 & latitude <=66.5)
s6<-volcano%>%filter(latitude >=66.5 & latitude <=90)

```

__Volcanos by Evidence Category__

Here is a bar chart showing how many volcanos have different types of known evidence of past eruptions. Here, we can see that most volcano eruptions have been observed. However, a sizable number have also been dated using various dating methods. Others, while not directly observed or dated, do have credible evidence. Comparatively, relatively few have uncertain eruption evidence.
```{r}
evidence_count<-volcano %>% 
  group_by(evidence_category) %>%
  summarise(no_rows4 = length(evidence_category))

ggplot(volcano, aes(x=evidence_category))+
  geom_bar()+
  ggtitle("Number of Volcanos by Evidence Category")+
  labs(x="Evidence Category", y="Number of Volcanos")+
  theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 45, hjust = 1, vjust = 0.95))+
  geom_text(data=evidence_count ,aes(x=evidence_category, y=no_rows4, label=no_rows4), nudge_y=30)+
   ggsave(file.path(figure_path, "volcano_count_by_evidence.pdf"))
```

__Eruptions Dataset__

```{r}
#simple histogram showing the number of eruptions of differing intesities
hist(eruptions$vei, main = "Volcano eruption index frequency", xlab = "Volcano eruption index", ylab = "Frequency")
```

```{r}
#interesting graphs showing the distribution of vei over longitude and latitude
#y axis is intensity and the x axis is longitude, size of dot = the number of observations at that given latitude
ggplot(eruptions, aes(x=latitude, y=vei))+
  geom_count(na.rm = T)+
  ggtitle("Volcano eruption index distributed over latitude")+
  labs(x="Latitude", y="vei")

ggplot(eruptions, aes(x=longitude, y=vei))+
  geom_count(na.rm = T)+
  ggtitle("Volcano eruption index distributed over longitude")+
  labs(x="Longitude", y="vei")
```

```{r}
#average number of years during which the eruption takes place
data <- eruptions[c(7,11)] %>%
  mutate(c = end_year - start_year)%>%
  drop_na()
mean(data$c)
```

```{r}
#bar plot showing the different methods used for evidence dating + a bar plot without the "Historical Observations" variable for easier viewing if the other methods
ggplot(eruptions, aes(x=evidence_method_dating))+
  geom_bar()+
  ggtitle("Prevalence of Different Eruption Dating Methods")+
  labs(x="Evidence Dating Methods", y="Number of Eruptions")+
  theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 40, hjust = 1, vjust = 1))

eruptions2 <- filter(eruptions, evidence_method_dating != "Historical Observations")
ggplot(eruptions2, aes(x=evidence_method_dating))+
  geom_bar()+
  ggtitle("Prevalence of Different Eruption Dating Methods")+
  labs(x="Evidence Dating Methods", y="Number of Eruptions")+
  theme(panel.grid=element_blank(), axis.text.x = element_text(angle = 40, hjust = 1, vjust = 1))
```
